An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-09-29 paper.tex
Update §5.3.2

Questions 2-4 relate to the \emph{power distance}
\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
individuals, which is the number of hierarchy levels between people who are
communicating. The focus is on the impact this has on people's voice or
lack thereof when the distance is large versus small. The conclusions from
this research area usually suggest that even small power distances have a
large effect of reducing the voice of the lower of the two levels. We
tackle this for the IETF in three different ways outlined in 
\S\ref{sec:org-dyn:hierarchy:rq2} -- \S\ref{sec:org-dyn:hierarchy:rq4}.

@@

@@

The effect the steepness of an organisation's hierarchy has on the
communication within is not well understood. This RQ is important because
once we know whether the IETF hierarchy is becoming steeper, we can look at
the effects this is having on the communication patterns of participants.

In this RQ, we look at the number of individuals and/or roles that inhabit
each hierarchy level. We also consider whether individuals in higher levels
are more active in the conversation and quantify the evolution of the
mailing list activity within each level of the IETF hierarchy. We find that
the middle level, Working Group Chairs (WGCs), is gaining in overall
proportion of activity, whilst the lowest level, RPs, is decreasing. The
top level, Area Directors (ADs), remains constant in proportion. The
increase in WGC activity coincides with an increase of WGCs per Working
Groups (WGs).

@@

For \textbf{RQ1}, to determine the steepness or diffusion of the hierarchy,
the proportion of nodes and activity are determined in each hierarchy
level. Figure~\ref{fig:activity_proportions}a shows the number of active
nodes in each level. The bottom level, RPs, are the largest contingent
consisting of 90--86\% of active participants; this proportion is
decreasing over time. The next level is WGCs which manage the WGs and are
6--10\% of active participants; this proportion is rising over time. The
top level is ADs which manage many WGs within an area and consist of less
than 1\% of the total active population of the IETF; this stays constant
over time. Therefore, the WGC level is becoming more diffused over time as
there are about 15 RPs for every one WGC in the organisation in 2013 versus
8 RPs per WGC in 2021.

The proportion of activity of nodes in the last year is also plotted and
split into hierarchy level in Figure~\ref{fig:activity_proportions}b. It is
clear that individuals at higher levels contribute a disproportionate share
to communication and this share is increasing over time, which is
consistent with findings in \cite{khare2022web}. Therefore, less of a
proportion of communication is flowing from the RPs at the same time as the
number of WGC roles is increasing. This further suggests that the IETF
hierarchy is becoming more diffused, in terms of mailing list
communication, which may make cooperation between hierarchy levels more
difficult.

\begin{figure}[t]
  \centering
  \small{a)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/active_node_proportions_year_sliding_window.png} \\
  \small{b)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/activity_proportions_year_sliding_window.png}
  \caption{
    Proportions of RPs, WGCs and ADs by a) number of active individuals and
    b) activity (number of emails sent to mailing lists). 
  }
  \label{fig:activity_proportions}
\end{figure}

Figure~\ref{fig:wgc_over_time}a shows the amount of nodes that inhabit the
WGC hierarchy level over our time period. There is a $35\%$ rise in the
amount of WGC roles whereas the amount of individuals that fill the WGC
roles remains mostly constant.

Figure~\ref{fig:wgc_over_time}b contains two estimates of the number of WGs
over time. The estimated number of WGs is likely the most accurate in the
middle 50\% of both plotted lines. Their discrepancy is the least (at most
10 WGs) from late 2014 to early 2019. The start of the COVID-19 pandemic
may also affect mailing list activity in 2020, and similar trends are
visible in the data in \cite{mcquistin2021characterising}. A full
explanation on how the data is estimated can be found in
section~\ref{sec:WG_activity}.

Figure~\ref{fig:wgc_over_time}c shows how many WGCs exist per WG, using
both estimates of the WG amount from Figure~\ref{fig:wgc_over_time}b. It is
clear there is a slow a rise in WGCs per WG over time. The ratio rises from
between 1.7-1.9 to 2.0-2.2 within the middle 50\%, late 2014 to early
2019.This perhaps represents a growing understanding in the IETF that it is
desirable to have two or more chairs per WG in case of
illness/unavailability or to avoid conflicts of interest. Therefore, this
plot again shows that the hierarchy is becoming more diffused, even if the
raw amount of individuals in the WGC roles has remained constant. The
multiple roles per individual may result in even more difficulty of RPs to
use their voice.

An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
omitted to save space as all three plots stay mostly constant throughout
the period. There are about 15 individual ADs split amongst 15 roles in 7-8
areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
while WGs are created and closed relatively frequently.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/over_time.png}
  \caption{
    Overview of WGC numbers and WG numbers over time. a) shows the number
    of WGC roles (some individuals hold multiple WGC roles) and the number
    of individuals who are WGCs b) shows two different estimates of the
    number of WGs (see section~\ref{sec:method}) and c) shows the mean
    number of WGCs per WG. Notice the non-zero-based y-axis exaggerates the
    variation in these figures.
  }
  \label{fig:wgc_over_time}
\end{figure}

An AD version of Figure~\ref{fig:wgc_over_time} is available, but it is
omitted to save space as all three plots stay mostly constant throughout
the period. There are about 15 individual ADs split amongst 15 roles in 7-8
areas with about 2 ADs per Area. The set of IETF Areas is broadly fixed,
while WGs are created and closed relatively frequently.

Table~\ref{tab:WGC_email} shows the emails sent and received for WGC and
ADs on mailing lists in the year before they first take that role, and the
year after. WGCs and ADs are both active in using the mailing lists, ADs
much more so. However, they don't become more active to any major degree
after taking the roles. This may suggest that the individuals who take on
higher hierarchy level roles are already contributing at a higher level.

@@

\pb{Summary}
In answer to RQ1 the picture is mixed, the number of individuals with WGC
roles has remained broadly constant but the number of WGC roles has
increased, indicating that individuals willing to become WGC have taken on
more such roles. The number of the higher level AD roles has, by design,
remained broadly constant over the time studied. The proportion of active
individuals with WGC roles has grown from 6\% to 10\% and the proportion of
the conversation taken by those roles has grown from 25\% to 35\% over the
period studied. Therefore the hierarchy has become more diffused, which may
mean RPs experience more difficulty in expressing their voice.

>>>>

Questions 2-4 relate to the \emph{power distance}
\cite{li2021does,duan2018authoritarian,guo2020inclusive} of
individuals, which is the number of hierarchy levels between people who are
communicating. The focus is on the impact this has on people's voice or
lack thereof when the distance is large versus small. The conclusions from
this research area usually suggest that even small power distances have a
large effect of reducing the voice of the lower of the two levels. We
tackle this for the IETF in three different ways in
\S\ref{sec:org-dyn:hierarchy:rq2} -- \S\ref{sec:org-dyn:hierarchy:rq4}.

@@

\todo{Do we need a separate methodology section?}

@@

% The effect the steepness of an organisation's hierarchy has on the
% communication within is not well understood. This RQ is important because
% once we know whether the IETF hierarchy is becoming steeper, we can look at
% the effects this is having on the communication patterns of participants.

In this section, we look at the number of individuals that inhabit each
level of the IETF hierarchy and their communication patterns, and quantify
the evolution of the mailing list activity within each level of the IETF.
We find that the middle level, Working Group Chairs, is gaining in overall
proportion of activity, whilst activity of participants at the lowest level
is decreasing. The top level of the hierarchy, area directors, remains
constant in proportion. The increase in working group chair activity
coincides with an increase in the number of chairs per working group.

@@

\pb{Role of Working Group Chairs:}
Figure \ref{fig:activity_proportions}(a) shows the number of participants
at each level of the hierarchy in IETF: Regular participants are the
largest contingent, comprising 90--86\% of active participants; this
proportion is decreasing over time.  The next level is Working Group
Chairs, comprising  6--10\% of active participants; this proportion is
rising and becoming more diffused over time, with about 15 regular
participants for every working group chair in the organisation in 2013
versus 8 participants per chair in 2021.  The top level is Area Directors
who consist of less than 1\% of the total active population of the IETF;
this stays constant over time.

The fraction of emails sent by participants at each level of the hierarchy
is shown in Figure~\ref{fig:activity_proportions}(b). Individuals at higher
levels contribute a disproportionate share to communication and this share
is increasing over time, consistent with \S\ref{sec:org-dyn:influence}. A
smaller proportion of email are sent by regular participants at the same
time as the number of working group chairs is increasing, further
suggesting that the IETF hierarchy is becoming more diffused, in terms of
mailing list communication, which may make cooperation between hierarchy
levels more difficult.

\begin{figure}[t]
  \centering
  \small{a)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/active_node_proportions_year_sliding_window.png} \\
  \small{b)} \\
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/activity_proportions_year_sliding_window.png}
  \caption{
    Proportion of regular participants, working group chairs, and area
    directors by (a) number of active individuals and (b) mailing list
    activity.
  }
  \label{fig:activity_proportions}
\end{figure}

%..................................................................................................
\pb{Number of Working Group Chairs:}
Figure~\ref{fig:wgc_over_time}(a) shows the number of working group chairs
our time. There is a $35\%$ rise in the number of chairs while the amount
of individuals that fill the chair roles remains mostly constant.

Figure~\ref{fig:wgc_over_time}(b) contains two estimates of the number of
working groups over time, based on tracking group events (i.e., Datatracker
metadata) and mailing list activity since groups may be recorded as being
active but have no activity, and vice-versa.  The discrepancy between the
two estimates is the least (at most 10 working groups) from late 2014 to
early 2019.

Figure~\ref{fig:wgc_over_time}(c) shows how many chairs exist per working
group, using both estimates of the number of working groups from
Figure~\ref{fig:wgc_over_time}(b). It is clear there is a slow a rise in
the number of chairs per working group over time. The ratio rises from
between 1.7-1.9 to 2.0-2.2 within the middle 50\%, late 2014 to early
2019.This perhaps represents a growing understanding in the IETF that it is
desirable to have two or more chairs per working group in case of
illness/unavailability or to avoid conflicts of interest. Therefore, this
plot again shows that the hierarchy is becoming more diffused, even if the
raw amount of individuals in the chair roles has remained constant. The
multiple roles per individual may result in even more difficulty of regular
participants to use their voice.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/over_time.png}
  \caption{
    Changes in the number of working groups and working group chairs over
    time: (a) shows the number of chairs and the number of individuals who
    are chairs; (b) shows two different estimates of the number of working
    groups; and c) shows the mean number of chairs per working group.
    Notice the non-zero-based y-axis exaggerates the variation in these
    figures.
  }
  \label{fig:wgc_over_time}
\end{figure}

%..................................................................................................
\pb{Number of Area Directors:}
A version of Figure~\ref{fig:wgc_over_time} considering area directors
rather than working group chairs is available, but omitted to save space as
all three plots stay mostly constant throughout the period. There are about
15 area directors split amongst 15 roles in 7-8 areas with typically 2 area
directors per Area. The set of IETF Areas changes slowly, while working
groups are created and closed relatively frequently.

%..................................................................................................
\pb{Activity prior to appointment:}
Table~\ref{tab:WGC_email} shows the amount of email sent and received for
working group chairs and area directors on mailing lists in the year before
they first take that role, and the year after. Chairs and area directors
are both active in using the mailing lists, area directors much more so.
However, they don't become more active to any major degree after taking the
roles. This may suggest that the individuals who take on higher hierarchy
level roles are already contributing at a higher level.

@@

%..................................................................................................
\pb{Summary}
The answer to the question of whether the higher levels of the hierarchy
growing or becoming more centralised, and whether they're associated with
an increased domination of the conversation, is mixed.  The number of
individuals with working group chair roles has remained broadly constant
but the number of such roles has increased, indicating that individuals
willing to become chairs have taken on more such roles.  The number of the
higher level area director roles has, by design, remained broadly constant
over the time studied. The proportion of active individuals with working
group chairs roles has grown from 6\% to 10\% and the proportion of the
conversation taken by those roles has grown from 25\% to 35\% over the
period studied. Therefore the hierarchy has become more diffused, which may
mean regular participants experience more difficulty in expressing their
voice.
